---
title: "Exercise R3.1: Querying a Public REST API"
subtitle: "STAT 540 — Week 3"
author: "Sokhna Amar"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will query a public REST API using R's `httr2` package, parse the JSON response, and perform basic analysis on the retrieved data.

## Task 1: Query the Open Meteo Weather API

Use the Open Meteo API (no authentication required) to fetch weather data for Fayetteville, NC:

```{r}
library(httr2)
library(dplyr)
library(tibble)

# Build the request
resp <- request("https://api.open-meteo.com/v1/forecast") |>
  req_url_query(
    latitude = 35.05,
    longitude = -78.88,
    daily = "temperature_2m_max,temperature_2m_min,precipitation_sum",
    timezone = "America/New_York",
    forecast_days = 7
  ) |>
  req_perform()

# Check status
cat("Status:", resp_status(resp), "\n")

# Parse JSON
weather <- resp |> resp_body_json()
str(weather, max.level = 2)
```

**Your turn:** What fields are in the response? Is the structure flat or nested?

> The fields contain variables with observations of weather data within the Eastern US. The structure is nested, with some observations containing specific information within it. 

## Task 2: Convert to a Tibble

```{r}
# Extract the daily data into a tibble
daily <- tibble(
  date = as.Date(unlist(weather$daily$time)),
  temp_max = unlist(weather$daily$temperature_2m_max),
  temp_min = unlist(weather$daily$temperature_2m_min),
  precip = unlist(weather$daily$precipitation_sum)
)

print(daily)
```

**Your turn:** What is the forecasted high temperature for tomorrow? What day has the most precipitation?

> The forecasted high temperature for tomorrow (9/9) is 30.1 degrees Celcius, and next Monday (9/14) is projected to have the most precipitation.

## Task 3: Query a Second API

Choose one of these APIs and query it:

- **REST Countries:** `https://restcountries.com/v3.1/name/{country}`
- **Universities:** `http://universities.hipolabs.com/search?country=United+States&name=fayetteville`
- **Open Trivia:** `https://opentdb.com/api.php?amount=10&category=18`

```{r}
library(httr2)

# Your code here — build a request, perform it, parse the response
resp <- request("https://opentdb.com/api.php?amount=10&category=18") |>
  req_url_query(amount = 10, category = 18) |>
  req_perform()

trivia <- resp |> resp_body_json()

str(trivia, max.level = 2)
trivia$response_code
length(trivia$results)
print(trivia$results)
```

**Your turn:** What API did you choose? How many records did you get? What did you learn about the data structure?

> I chose the open trivia API. I got ten records of trivia questions, of which contained the question, the answer, and category.

## Task 4: Error Handling

Deliberately cause an error (wrong URL, bad parameter) and handle it gracefully:

```{r}
library(httr2)

# Try a request that will fail
resp <- request("https://api.github.com/users/THIS_USER_DOES_NOT_EXIST_12345") |>
  req_error(is_error = function(resp) FALSE) |>  # Don't stop on error
  req_perform()

status <- resp_status(resp)
cat("Status code:", status, "\n")

if (status == 200) {
  cat("Success!\n")
} else if (status == 404) {
  cat("Not found — check the endpoint\n")
} else {
  cat("Unexpected error:", status, "\n")
}
```

## Submission

```bash
git add week03/exercises/R3.1*
git commit -m "Complete Exercise R3.1: API queries with httr2"
git push origin main
```
