---
title: "Exercise R3.2: Web Scraping with rvest"
subtitle: "STAT 540 — Week 3"
author: "Your Name Here"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will scrape data from web pages using R's `rvest` package, working with both HTML tables and individual elements.

## Task 1: Scrape a Wikipedia Table

```{r}
library(rvest)
library(dplyr)

url <- "https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_population"
page <- read_html(url)

# Extract all tables
tables <- page |> html_table()
cat("Number of tables found:", length(tables), "\n")

# Get the main population table (usually the first or second)
states <- tables[[1]]
glimpse(states)
```

**Your turn:** How many rows and columns does the table have? What cleaning would it need before analysis (e.g., column names, data types)?

> *[Your answer]*

## Task 2: Clean the Scraped Table

```{r}
library(janitor)
library(readr)

# Your cleaning code here:
# - Use clean_names() for consistent column names
# - Convert population columns from character to numeric
# - Remove footnote markers or special characters
# states_clean <- states |> ...
```

## Task 3: Scrape books.toscrape.com

This site is designed for scraping practice:

```{r}
library(rvest)
library(tibble)

page <- read_html("https://books.toscrape.com")

# Extract titles
titles <- page |> html_elements("h3 a") |> html_attr("title")

# Extract prices
prices <- page |> html_elements(".price_color") |> html_text()

# Extract star ratings
ratings <- page |> html_elements("p.star-rating") |> html_attr("class")

# Combine
books <- tibble(
  title = titles,
  price = prices,
  rating_class = ratings
)
print(books)
```

**Your turn:** How would you convert the price text (e.g., "£51.77") to a numeric value? How would you extract just the rating word (e.g., "Three") from the rating class?

> *[Your answer]*

## Task 4: Check robots.txt

```{r}
library(httr2)

# Check what scraping is allowed
robots <- request("https://books.toscrape.com/robots.txt") |>
  req_perform() |>
  resp_body_string()

cat(robots)
```

**Your turn:** Does robots.txt allow scraping this site? What does the content say?

> *[Your answer]*

## Task 5: Ethical Reflection

Write 100 words on the ethics of web scraping. Address: when is it appropriate, what precautions should you take, and why does robots.txt matter?

> *[Your reflection]*

## Submission

```bash
git add week03/exercises/R3.2*
git commit -m "Complete Exercise R3.2: Web scraping with rvest"
git push origin main
```
