---
title: "Exercise R1.2: First Reproducible R Script"
subtitle: "STAT 540 — Week 1"
author: "Sokhna Amar"
date: last-modified
execute:
  enabled: true
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will create a reproducible R analysis using a built-in dataset, practice basic data inspection, and commit your work to GitHub.

## Task 1: Load and Inspect mtcars

The `mtcars` dataset is one of R's classic built-in datasets. It contains data from the 1974 *Motor Trend* magazine about 32 automobiles.

```{r}
# Load the dataset (built-in, no package needed)
data(mtcars)

# View the first 5 rows
head(mtcars, 5)

# View the last 3 rows
tail(mtcars, 3)

# What are the dimensions?
dim(mtcars)

# What are the column names?
names(mtcars)

# Structure of the dataset
str(mtcars)
```

**Your turn:** In your own words, describe what each row represents and name three variables you find interesting.

> Each row represents the individual car models in the mtcars dataset, with specific details according to the variables. I find the 'mpg', 'wt', and 'hp' variables to be quite interesting, as these are the details I pay attention to the most when looking into cars. 

## Task 2: Compute Summary Statistics

```{r}
# Mean miles per gallon
mean_mpg <- mean(mtcars$mpg)
cat("Mean MPG:", mean_mpg, "\n")

# Median horsepower
median_hp <- median(mtcars$hp)
cat("Median HP:", median_hp, "\n")

# Standard deviation of weight
sd_wt <- sd(mtcars$wt)
cat("SD of Weight:", sd_wt, "\n")

# Range of quarter-mile time
range_qsec <- range(mtcars$qsec)
cat("Quarter-mile time range:", range_qsec[1], "to", range_qsec[2], "\n")
```

**Your turn:** Compute the following and report the values:

1. The mean weight (`wt`) of all cars: ___
2. The maximum horsepower (`hp`): ___
3. The number of cars with 4 cylinders: ___

```{r}
# Your code for the three computations above:
# 1.
mean(mtcars$wt)

# 2.
max(mtcars$hp)

# 3.
sum(mtcars$cyl == 4)
```

The mean weight of the cars is 3217.25 pounds. The maximum horsepower is 335, and there are 11 cars with 4 cylinders. 

## Task 3: Grouped Summaries

Compute the mean MPG for each number of cylinders:

```{r}
# Base R approach
tapply(mtcars$mpg, mtcars$cyl, mean)

# Tidyverse approach
library(dplyr)
mtcars |>
  group_by(cyl) |>
  summarize(
    mean_mpg = mean(mpg),
    mean_hp = mean(hp),
    count = n()
  )
```

**Your turn:** Write a sentence interpreting the relationship between number of cylinders and fuel efficiency:

> There is a negative correlation between the number of cylinders and fuel efficiency; the more cylinders there are, the less mpg the cars have. 

## Task 4: Your First Visualization

```{r}
library(ggplot2)

# Scatterplot of weight vs. MPG, colored by cylinder count
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(
    title = "Car Weight vs. Fuel Efficiency",
    subtitle = "Data from 1974 Motor Trend magazine",
    x = "Weight (1,000 lbs)",
    y = "Miles Per Gallon",
    color = "Cylinders"
  ) +
  theme_minimal()
```

**Your turn:** Modify the plot above to also include a smooth trend line for each cylinder group. Hint: add `geom_smooth(method = "lm", se = FALSE)` to the plot.

```{r}
# Your modified plot code here:
ggplot(mtcars, aes(x = wt, y = mpg, color = factor(cyl))) +
  geom_point(size = 3) +
  labs(
    title = "Car Weight vs. Fuel Efficiency",
    subtitle = "Data from 1974 Motor Trend magazine",
    x = "Weight (1,000 lbs)",
    y = "Miles Per Gallon",
    color = "Cylinders"
  ) +
  theme_minimal() +
  geom_smooth(method = "lm", se = FALSE)

```

## Task 5: Write a Data Summary

In 3–5 sentences, summarize what you learned about the mtcars dataset. Include at least one specific number from your analysis and one observation from your visualization.

> The mt cars dataset contains information about fuel consumption, horsepower, and various design characteristics from various cars from 1973-74. What I found particularly interesting was the correlation between the number of cylinders, which ranged from 4-8, and the vehicle's average fuel efficiency, which ranged from a mean of about 15 mpg to 26 mpg. To be more specific, the lower the amount of cylinders, the farther a car could travel on a gallon of gas. Visualizing this on a scatter plot also showed how weight correlated with the number of cylinders as well. 

## Task 6: Commit to GitHub

1. Save this completed `.qmd` file
2. Render it to HTML
3. In the terminal, commit and push:

```bash
git add week01/exercises/R1.2-first-script.qmd
git add week01/exercises/R1.2-first-script.html
git commit -m "Complete Exercise R1.2: mtcars analysis"
git push origin main
```

## Submission Checklist

- [ ] All code chunks run without errors
- [ ] All "Your turn" sections are completed
- [ ] Data summary paragraph is written
- [ ] Modified visualization includes trend lines
- [ ] Both `.qmd` and `.html` files are committed to GitHub
- [ ] Commit message is descriptive
