---
title: "Exercise R5.1: Feature Engineering & Recipes Pipeline"
subtitle: "STAT 540 — Week 5"
author: "Sokhna Amar"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will engineer date-based features and build a complete preprocessing pipeline using the recipes package.

## Task 1: Date Feature Engineering with lubridate

Using the flights dataset, extract temporal features:

```{r}
library(lubridate)
library(dplyr)
library(nycflights13)

flights_feat <- flights |>
  mutate(
    date = make_date(year, month, day),
    dow = wday(date, label = TRUE),
    is_weekend = dow %in% c("Sat", "Sun"),
    week = isoweek(date),
    is_summer = month %in% 6:8
  )

# How do delays differ by day of week?
flights_feat |>
  filter(!is.na(arr_delay)) |>
  group_by(dow) |>
  summarize(avg_delay = mean(arr_delay), pct_late = mean(arr_delay > 0) * 100)
```

**Your turn:** Which day of the week has the highest average delay? Does the weekend pattern match your intuition?

> Thursday has the highest average delay. I was suprised about the weekend pattern, though. I thought it would be higher.

## Task 2: Build a Recipes Pipeline

Build a preprocessing recipe for the Ames Housing dataset (or another dataset of your choice):

```{r}
library(recipes)
library(dplyr)

# If using a built-in dataset:
data(ames, package = "modeldata")  # Or use diamonds

rec <- recipe(Sale_Price ~ ., data = ames) |>
  step_impute_median(all_numeric_predictors()) |>
  step_impute_mode(all_nominal_predictors()) |>
  step_log(Sale_Price, base = 10) |>         # Log-transform the outcome
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_nzv(all_predictors())

# Prep and bake
rec_prepped <- prep(rec)
processed <- bake(rec_prepped, new_data = ames)

# Compare before/after
cat("Before:", ncol(ames), "columns\n")
cat("After:", ncol(processed), "columns\n")
```

**Your turn:** Why did the column count change? What did `step_dummy()` and `step_nzv()` do?

> The column count changed because `step_dummy()` and `step_nzv()` pull in opposite directions. `step_dummy()` converts each nominal (categorical) predictor into a set of 0/1 dummy variables — one fewer than the number of levels it has, using one level as the reference — so a single column like Neighborhood becomes many numeric columns, which sharply increases the total. `step_nzv()` then removes near-zero-variance predictors — columns (often those new dummy variables) that are almost constant, i.e. one value dominates almost all the rows — since these carry little useful information for modeling. The net effect on column count is the balance between dummy-encoding adding many columns and `step_nzv()` removing the low-information ones.

## Task 3: Document Your Pipeline

For each step in your recipe, explain WHY you included it:

| Step | What It Does | Why You Included It |
|------|-------------|-------------------|
| `step_impute_median()` | Replaces missing values in numeric predictors with the median of that column. | Median is robust to outliers and skew (unlike the mean), and many models can't handle `NA`s directly; imputing lets us keep rows instead of dropping them. |
| `step_impute_mode()` | Replaces missing values in categorical (nominal) predictors with the most frequent category. | Mode is the natural central tendency for categorical data (you can't take a mean/median of text labels), so it's the standard way to fill in missing categories without discarding data. |
| `step_log()` | Applies a log transform (base 10 here) to `Sale_Price`. | Home prices are typically right-skewed (a few very expensive homes stretch the distribution). Log-transforming the outcome makes its distribution more symmetric/normal, which helps linear models satisfy their assumptions and often improves prediction accuracy. |
| `step_normalize()` | Centers and scales all numeric predictors to mean 0, standard deviation 1. | Predictors are on very different scales (e.g., square footage vs. number of bathrooms). Normalizing puts them on comparable footing, which matters for models sensitive to scale (e.g., regularized regression, KNN, PCA) and makes coefficients easier to compare. |
| `step_dummy()` | Converts nominal predictors into a set of 0/1 dummy variables (k−1 per categorical column). | Most models require numeric input and can't use text/factor categories directly. Dummy encoding represents categorical information numerically without implying a false ordinal relationship between categories. |
| `step_nzv()` | Removes predictors with near-zero variance (i.e., columns that are almost constant across observations). | Near-constant predictors (often rare dummy variables created by `step_dummy()`) carry almost no useful information for modeling, can cause numerical instability, and just add noise/computation cost — so it's efficient to drop them. |

## Submission

```bash
git add week05/exercises/R5.1*
git commit -m "Complete Exercise R5.1: Feature engineering and recipes pipeline"
git push origin main
```
