---
title: "Exercise RP5.2: Creative Feature Engineering"
subtitle: "STAT 540 — Week 5"
author: "Your Name Here"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will apply creative feature engineering to a real dataset. The goal is to think like a domain expert — not just apply transformations mechanically.

## The Scenario

You work for an e-commerce company. Given a dataset of customer transactions, your task is to engineer features that could help predict **customer churn** (whether a customer stops buying).

You may use the Online Retail dataset from UCI, a similar e-commerce dataset, or simulate one.

## Task 1: Load and Inspect

```{r}
# R version (or use Python — your choice)
# Download from UCI or use a provided sample
library(readr)
library(dplyr)

# retail <- read_csv("online_retail.csv")
# glimpse(retail)
```

## Task 2: Engineer 5 Features

Create at least 5 engineered features. For each, explain your **hypothesis** — why you believe this feature would help predict churn.

| # | Feature Name | How Computed | Hypothesis |
|---|-------------|-------------|------------|
| 1 | *[name]* | *[computation]* | *[why it matters]* |
| 2 | *[name]* | *[computation]* | *[why it matters]* |
| 3 | *[name]* | *[computation]* | *[why it matters]* |
| 4 | *[name]* | *[computation]* | *[why it matters]* |
| 5 | *[name]* | *[computation]* | *[why it matters]* |

```{r}
# Your feature engineering code here:

```

## Task 3: Before/After Summary

Show summary statistics for your engineered features:

```{r}
# Your code to show summary stats of the new features:

```

## Task 4: Rank Your Features

Which of your 5 features do you expect to be most useful for predicting churn? Rank them and explain your reasoning (1 = most useful):

1. *[Feature]* — *[Why]*
2. *[Feature]* — *[Why]*
3. *[Feature]* — *[Why]*
4. *[Feature]* — *[Why]*
5. *[Feature]* — *[Why]*

## Task 5: Reflection

In 100 words, reflect: How did domain thinking (understanding the business context) change what features you created compared to purely mechanical transformations?

> *[Your reflection]*

## Submission Checklist

- [ ] Dataset loaded and inspected
- [ ] 5 features engineered with hypotheses
- [ ] Code runs and produces the features
- [ ] Summary statistics shown
- [ ] Features ranked with reasoning
- [ ] Reflection written
- [ ] Committed to GitHub
