---
title: "Exercise RP5.2: Creative Feature Engineering"
subtitle: "STAT 540 — Week 5"
author: "Sokhna Amar"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will apply creative feature engineering to a real dataset. The goal is to think like a domain expert — not just apply transformations mechanically.

## The Scenario

You work for an e-commerce company. Given a dataset of customer transactions, your task is to engineer features that could help predict **customer churn** (whether a customer stops buying).

You may use the Online Retail dataset from UCI, a similar e-commerce dataset, or simulate one.

## Task 1: Load and Inspect

```{r}
library(readr)
library(dplyr)
library(purrr)

# Simulating a transactions dataset rather than downloading UCI Online Retail,
# so the exercise is self-contained and reproducible.
set.seed(540)
n_customers <- 300
reference_date <- as.Date("2024-01-01")

retail <- map_dfr(1:n_customers, function(id) {
  n_orders <- sample(1:20, 1)
  order_dates <- sort(reference_date - sample(1:730, n_orders, replace = TRUE))
  tibble(
    customer_id = id,
    order_date = order_dates,
    order_amount = round(rlnorm(n_orders, meanlog = 3.5, sdlog = 0.6), 2)
  )
})

glimpse(retail)
```

## Task 2: Engineer 5 Features

Create at least 5 engineered features. For each, explain your **hypothesis** — why you believe this feature would help predict churn.

| # | Feature Name | How Computed | Hypothesis |
|---|-------------|-------------|------------|
| 1 | `recency_days` | Days between `reference_date` and the customer's most recent order | Supposedly, customers who haven't bought in a long time are the most likely to have already churned |
| 2 | `frequency` | Count of orders per customer | Habitual, frequent buyers are more embedded in the relationship and less likely to leave than one-time or occasional shoppers |
| 3 | `avg_order_value` | Mean `order_amount` per customer | Distinguishes low-value browsers from high-value customers, which is useful for prioritizing *which* churn is worth acting on, not just predicting it |
| 4 | `purchase_interval_cv` | Coefficient of variation (sd/mean) of the gaps in days between a customer's consecutive orders | Captures rhythm, not just volume; a customer with an erratic, unpredictable buying cadence (high CV) may be more disengaged than one with a steady, predictable habit, even at the same frequency |
| 5 | `spend_trend` | % change in average order value between a customer's earlier orders and their more recent orders | A customer whose spending is declining relative to their *own* history is showing early churn risk before recency alone would flag it |

```{r}
# Your feature engineering code here:
features <- retail |>
  arrange(customer_id, order_date) |>
  group_by(customer_id) |>
  summarize(
    recency_days = as.numeric(reference_date - max(order_date)),
    frequency = n(),
    avg_order_value = mean(order_amount),
    purchase_interval_cv = {
      gaps <- as.numeric(diff(order_date))
      if (length(gaps) >= 2) sd(gaps) / mean(gaps) else NA_real_
    },
    spend_trend = {
      half <- ceiling(n() / 2)
      early <- mean(order_amount[1:half])
      late  <- mean(order_amount[(half + 1):n()])
      if (n() >= 3) (late - early) / early * 100 else NA_real_
    },
    .groups = "drop"
  )

glimpse(features)
```

## Task 3: Before/After Summary

Show summary statistics for your engineered features:

```{r}
# Your code to show summary stats of the new features:
library(skimr)

skim(features)
```

`purchase_interval_cv` and `spend_trend` will show `NA` for customers with fewer than 3 orders — that's expected, since there aren't enough transactions to measure a gap pattern or a before/after trend for them, not a bug.

## Task 4: Rank Your Features

Which of your 5 features do you expect to be most useful for predicting churn? Rank them and explain your reasoning (1 = most useful):

1. `recency_days` — Thus one is the single strongest classic churn signal; a customer who hasn't purchased in months is already showing the behavior we're trying to predict, almost by definition.
2. `spend_trend` — A leading indicator rather than a lagging one: it can flag a customer sliding toward churn *before* their recency gets bad enough to notice, by catching the decline in their own spending pattern.
3. `frequency` — A solid engagement measure, but it's an aggregate over the customer's whole history, so it reacts slowly to recent changes compared to recency or trend.
4. `purchase_interval_cv` — Genuinely informative about behavioral consistency, but noisier and undefined for low-order customers, which limits how broadly it applies.
5. `avg_order_value` — Least directly tied to churn itself — a big spender can still leave — so it's more useful for prioritizing *which* churn matters than for predicting churn in the first place.

## Task 5: Reflection

In 100 words, reflect: How did domain thinking (understanding the business context) change what features you created compared to purely mechanical transformations?

> Purely mechanical feature engineering would have stopped at aggregates such as total spend, order count, average value, treating each customer as a static summary. Thinking about *why* a customer churns pushed me toward features that capture change over time instead: `spend_trend` asks whether someone is pulling back relative to their own past, and `purchase_interval_cv` asks whether their buying rhythm is destabilizing. Neither is a transformation you'd stumble into mechanically; they came from imagining what an actual disengaging customer's data would look like before they fully stop buying, not just describing what they've already done.

## Submission Checklist

- [x] Dataset loaded and inspected
- [x] 5 features engineered with hypotheses
- [x] Code runs and produces the features
- [x] Summary statistics shown
- [x] Features ranked with reasoning
- [x] Reflection written
- [x] Committed to GitHub
