---
title: "Exercise R2.2: Working with Tibbles"
subtitle: "STAT 540 — Week 2"
author: "Your Name Here"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will explore the differences between base R data frames and tibbles, practice data inspection techniques, and build the habit of systematic dataset exploration.

## Task 1: Data Frame vs Tibble

Convert the built-in `mtcars` data frame to a tibble and observe the differences:

```{r}
library(tibble)
library(dplyr)

# Base R data frame
print(mtcars)    # Notice: prints ALL rows

# Convert to tibble
mtcars_tbl <- as_tibble(mtcars, rownames = "car")
print(mtcars_tbl)   # Notice: prints first 10 rows, shows column types
```

**Your turn:** List three differences you observe between printing a data frame and printing a tibble:

1. *[Difference 1]*
2. *[Difference 2]*
3. *[Difference 3]*

## Task 2: Inspecting with glimpse()

```{r}
library(dplyr)

# glimpse is the tidyverse equivalent of str()
glimpse(mtcars_tbl)

# Compare to base R
str(mtcars)
```

**Your turn:** What information does `glimpse()` show that `print()` does not?

> *[Your answer]*

## Task 3: Using skimr for Comprehensive Summaries

```{r}
library(skimr)

# Full dataset skim
skim(mtcars_tbl)

# Grouped skim
mtcars_tbl |>
  group_by(cyl) |>
  skim(mpg, hp, wt)
```

**Your turn:** From the `skim()` output, answer:

1. Which numeric variable has the highest coefficient of variation (SD / mean)? ___
2. Are there any missing values in mtcars? ___
3. How does mean `mpg` differ across cylinder groups? ___

## Task 4: Subsetting Behavior

An important tibble safety feature — subsetting always returns a tibble:

```{r}
# Data frame: single column returns a vector
mtcars[, "mpg"]         # vector
class(mtcars[, "mpg"])  # "numeric"

# Tibble: single column returns a tibble
mtcars_tbl[, "mpg"]           # still a tibble
class(mtcars_tbl[, "mpg"])    # "tbl_df"

# To extract as a vector from a tibble, use $ or [[
mtcars_tbl$mpg               # vector
mtcars_tbl[["mpg"]]          # vector
```

**Your turn:** Why is the tibble behavior safer for programming? Think of a situation where the data frame behavior could cause a bug.

> *[Your answer]*

## Task 5: Build Your Own Tibble

Create a tibble from scratch representing 5 students in the MIDS program:

```{r}
library(tibble)

# Your turn: create a tibble with at least 4 columns
# Include: name (character), age (numeric), concentration (factor or character),
# and one column of your choice

mids_students <- tibble(
  # Your code here
)

glimpse(mids_students)
```

## Task 6: Inspect a Real Dataset

Choose one of the built-in datasets below, convert to tibble if needed, and perform a complete inspection:

- `nycflights13::flights`
- `ggplot2::diamonds`
- `palmerpenguins::penguins`

```{r}
# Your choice — load and inspect
library(palmerpenguins)  # or another package

df <- as_tibble(penguins)

# Run your inspection checklist:
# 1. Shape
dim(df)

# 2. Types
glimpse(df)

# 3. Head/Tail
head(df)
tail(df)

# 4. Missing values
colSums(is.na(df))

# 5. Summary
summary(df)

# 6. Comprehensive
library(skimr)
skim(df)
```

**Your turn:** Write a 3–5 sentence inspection summary of your chosen dataset:

> *[Your summary]*

## Submission Checklist

- [ ] Data frame vs tibble differences documented
- [ ] glimpse() vs print() comparison explained
- [ ] skimr questions answered with specific numbers
- [ ] Subsetting safety question addressed
- [ ] Custom tibble created
- [ ] Real dataset inspected with full checklist
- [ ] Files committed to GitHub
