---
title: "Exercise R2.2: Working with Tibbles"
subtitle: "STAT 540 — Week 2"
author: "Sokhna Amar"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will explore the differences between base R data frames and tibbles, practice data inspection techniques, and build the habit of systematic dataset exploration.

## Task 1: Data Frame vs Tibble

Convert the built-in `mtcars` data frame to a tibble and observe the differences:

```{r}
library(tibble)
library(dplyr)

# Base R data frame
print(mtcars)    # Notice: prints ALL rows

# Convert to tibble
mtcars_tbl <- as_tibble(mtcars, rownames = "car")
print(mtcars_tbl)   # Notice: prints first 10 rows, shows column types
```

**Your turn:** List three differences you observe between printing a data frame and printing a tibble:

1. *I can see that tibbles, when printed out, have a more explicit view of the different data types for each of the columns, or variables.*
2. *Each of the rows, or observations, are numbered.*
3. *There is a clear view of how big the data structure is.*

## Task 2: Inspecting with glimpse()

```{r}
library(dplyr)

# glimpse is the tidyverse equivalent of str()
glimpse(mtcars_tbl)

# Compare to base R
str(mtcars)
```

**Your turn:** What information does `glimpse()` show that `print()` does not?

> *`glimpse()` prints out the tibbles differently than `print()` does--the difference lies mostly in the layout of the rows and columns. The former has a more transposed view of the layout, so that the columns run vertically down the edge, with each of the values next to it shown horizontally, while the opposite is true for the latter.*

## Task 3: Using skimr for Comprehensive Summaries

```{r}
library(skimr)

# Full dataset skim
skim(mtcars_tbl)

# Grouped skim
mtcars_tbl |>
  group_by(cyl) |>
  skim(mpg, hp, wt)
```

**Your turn:** From the `skim()` output, answer:

1. Which numeric variable has the highest coefficient of variation (SD / mean)? ___
2. Are there any missing values in mtcars? ___
3. How does mean `mpg` differ across cylinder groups? ___

> 1. The largest coefficient of variation is for am at around 1.228.
2. There are no missing variables. 
3. Across the different variable groups: 4 cylinders: 26.66 mpg
6 cylinders: 19.74 mpg
8 cylinders: 15.10 mpg

## Task 4: Subsetting Behavior

An important tibble safety feature — subsetting always returns a tibble:

```{r}
# Data frame: single column returns a vector
mtcars[, "mpg"]         # vector
class(mtcars[, "mpg"])  # "numeric"

# Tibble: single column returns a tibble
mtcars_tbl[, "mpg"]           # still a tibble
class(mtcars_tbl[, "mpg"])    # "tbl_df"

# To extract as a vector from a tibble, use $ or [[
mtcars_tbl$mpg               # vector
mtcars_tbl[["mpg"]]          # vector
```

**Your turn:** Why is the tibble behavior safer for programming? Think of a situation where the data frame behavior could cause a bug.

> Tibbles are better for tabular data because columns are still treated as tibbles, so the structure of the data is respected. One scenario where using a data frame instead might be when a specific column is selected which turns it into a vector, which makes the code buggy. Tibbles make subsetting more predictable. 

## Task 5: Build Your Own Tibble

Create a tibble from scratch representing 5 students in the MIDS program:

```{r}
library(tibble)

# Your turn: create a tibble with at least 4 columns
# Include: name (character), age (numeric), concentration (factor or character),
# and one column of your choice

mids_students <- tibble(
  name = c("Amina", "Ben", "Carlos", "Diana", "Ethan"),
  age = c(24, 27, 31, 26, 29),
  concentration = c("Data Science", "Statistics", "Data Science", "Biostatistics", "Statistics"),
  prior_experience = c(TRUE, FALSE, TRUE, TRUE, FALSE)
)

glimpse(mids_students)
```

## Task 6: Inspect a Real Dataset

Choose one of the built-in datasets below, convert to tibble if needed, and perform a complete inspection:

- `nycflights13::flights`
- `ggplot2::diamonds`
- `palmerpenguins::penguins`

```{r}
# Your choice — load and inspect
library(palmerpenguins)  # or another package

df <- as_tibble(penguins)

# Run your inspection checklist:
# 1. Shape
dim(df)

# 2. Types
glimpse(df)

# 3. Head/Tail
head(df)
tail(df)

# 4. Missing values
colSums(is.na(df))

# 5. Summary
summary(df)

# 6. Comprehensive
library(skimr)
skim(df)
```

**Your turn:** Write a 3–5 sentence inspection summary of your chosen dataset:

> The penguins dataset contains 344 observations and 8 variables describing penguin species, island, physical measurements, sex, and year. It includes three species: Adelie, Chinstrap, and Gentoo. The dataset contains some missing values, especially in the sex variable, with a few missing physical measurements as well. Gentoo penguins generally have larger body mass and longer flippers than Adelie and Chinstrap penguins.


## Submission Checklist

- [x] Data frame vs tibble differences documented
- [x] glimpse() vs print() comparison explained
- [x] skimr questions answered with specific numbers
- [x] Subsetting safety question addressed
- [x] Custom tibble created
- [x] Real dataset inspected with full checklist
- [x] Files committed to GitHub
