---
title: "Exercise R4.1: Five-Verb dplyr Pipeline"
subtitle: "STAT 540 — Week 4"
author: "Your Name Here"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will build a complete dplyr pipeline using the nycflights13 dataset, practicing all five core verbs plus grouping.

## Task 1: Delayed Flights by Carrier

Build a single piped expression that:

1. Filters to flights with arrival delay greater than 0
2. Groups by carrier
3. Computes: average delay, median delay, number of flights, and percent more than 1 hour late
4. Sorts by average delay descending

```{r}
library(nycflights13)
library(dplyr)

# Your pipeline here:
flights |>
  filter(arr_delay > 0) |>
  group_by(carrier) |>
  summarize(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    median_delay = median(arr_delay, na.rm = TRUE),
    n = n(),
    pct_over_60 = mean(arr_delay > 60, na.rm = TRUE) * 100
  ) |>
  arrange(desc(avg_delay))
```

**Your turn:** Which carrier has the highest average delay? Which has the most flights but is still late frequently?

> *[Your answer]*

## Task 2: Custom Analysis Question

Write your own analysis question about the flights data, then answer it with a dplyr pipeline. Your pipeline must use at least 4 of the 5 verbs (filter, select, mutate, arrange, summarize).

**My question:** Among routes (origin-destination pairs) flown at least 100 times in 2013, which have the best on-time performance, as measured by lowest average arrival delay?

```{r}
# Your pipeline here:
flights |>
  filter(!is.na(arr_delay)) |>
  select(origin, dest, arr_delay) |>
  mutate(route = paste(origin, dest, sep = "-")) |>
  group_by(route) |>
  summarize(
    avg_arr_delay = mean(arr_delay),
    n = n()
  ) |>
  filter(n >= 100) |>
  arrange(avg_arr_delay)
```

**Your answer:** As seen with the results, most of the flights with the best on-time performance tend to land early, and are long haul flights, such as  EWR-SNA (Newark→Orange County CA), JFK-HNL (Honolulu), JFK-STT (St. Thomas USVI), which suggests that airlines tend to pad longer flights with extra time on schedule, making it more likely they arrive earlier than their scheduled time.  

## Task 3: Reshape World Bank-Style Data

```{r}
library(tidyr)

# Simulated wide data
gdp_wide <- tibble(
  country = c("USA", "China", "India", "Brazil", "Germany"),
  `2020` = c(21.0, 14.7, 2.7, 1.4, 3.8),
  `2021` = c(23.3, 17.7, 3.2, 1.6, 4.2),
  `2022` = c(25.4, 17.9, 3.4, 1.9, 4.1),
  `2023` = c(26.9, 18.5, 3.7, 2.1, 4.5)
)

# Pivot to long format
gdp_long <- gdp_wide |>
  pivot_longer(cols = -country, names_to = "year", values_to = "gdp_trillion") |>
  mutate(year = as.integer(year))

print(gdp_long)
```

**Your turn:** Which country had the highest GDP growth (absolute change) from 2020 to 2023? Write a pipeline to compute this.

```{r}
# Your code here:
gdp_long |>
  filter(year %in% c(2020, 2023)) |>
  pivot_wider(names_from = year, values_from = gdp_trillion, names_prefix = "gdp_") |>
  mutate(growth = gdp_2023 - gdp_2020) |>
  arrange(desc(growth))
```

## Submission

```bash
git add week04/exercises/R4.1*
git commit -m "Complete Exercise R4.1: dplyr pipeline with flights data"
git push origin main
```
