---
title: "Exercise RP4.2: Wrangling Challenge (R or Python)"
subtitle: "STAT 540 — Week 4"
author: "Your Name Here"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will take a messy dataset and produce a clean, tidy version. You may use R or Python (or both). The focus is on **documenting your reasoning** — not just writing code.

## The Messy Dataset

Download the messy dataset provided by the instructor (or use the Palmer Penguins dataset with intentionally introduced messiness).

```{r}
library(palmerpenguins)
library(dplyr)
library(tidyr)

# Start with penguins — it already has some real-world messiness
df <- penguins
glimpse(df)

# How messy is it?
cat("Missing values per column:\n")
colSums(is.na(df))
```

## Task 1: Data Quality Assessment

Before cleaning, assess the current state:

1. How many rows and columns?
2. Which columns have missing values? What percentage?
3. Are any columns the wrong type?
4. Are there any obvious outliers or impossible values?

> *[Your assessment]*

## Task 2: Cleaning Pipeline

Build a step-by-step cleaning pipeline. **Comment every transformation** explaining WHY you made that choice:

```{r}
# Example structure — modify for your data:
df_clean <- df |>
  # Step 1: [Why]
  # ... |>
  # Step 2: [Why]
  # ... |>
  # Step 3: [Why]
  # ...
  identity()   # Placeholder — remove this

glimpse(df_clean)
```

## Task 3: Before/After Comparison

Show summary statistics before and after cleaning:

```{r}
library(skimr)

cat("=== BEFORE ===\n")
skim(df)

cat("\n=== AFTER ===\n")
skim(df_clean)
```

**Your turn:** What changed? What improved? Did you lose any data, and was that loss justified?

> *[Your answer]*

## Task 4: Document Your Decisions

For each cleaning step, complete this table:

| Step | What You Did | Why | What You Considered But Rejected |
|------|-------------|-----|--------------------------------|
| 1 | *[action]* | *[reason]* | *[alternative]* |
| 2 | *[action]* | *[reason]* | *[alternative]* |
| 3 | *[action]* | *[reason]* | *[alternative]* |

## Submission Checklist

- [ ] Data quality assessment completed
- [ ] Cleaning pipeline with comments explaining every step
- [ ] Before/after summary statistics shown
- [ ] Decision documentation table filled in
- [ ] Committed to GitHub

```bash
git add week04/exercises/RP4.2*
git commit -m "Complete Exercise RP4.2: Wrangling challenge with documented decisions"
git push origin main
```
