---
title: "Exercise RP4.2: Wrangling Challenge (R or Python)"
subtitle: "STAT 540 — Week 4"
author: "Sokhna Amar"
date: last-modified
execute:
  enabled: false
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    code-fold: false
---

## Overview

In this exercise, you will take a messy dataset and produce a clean, tidy version. You may use R or Python (or both). The focus is on **documenting your reasoning** — not just writing code.

## The Messy Dataset

Download the messy dataset provided by the instructor (or use the Palmer Penguins dataset with intentionally introduced messiness).

```{r}
library(palmerpenguins)
library(dplyr)
library(tidyr)

# Start with penguins — it already has some real-world messiness
df <- penguins
glimpse(df)

# How messy is it?
cat("Missing values per column:\n")
colSums(is.na(df))
```

## Task 1: Data Quality Assessment

Before cleaning, assess the current state:

1. How many rows and columns?
2. Which columns have missing values? What percentage?
3. Are any columns the wrong type?
4. Are there any obvious outliers or impossible values?

> 1. There are 344 rows and 8 columns. 
2. The bill length, bill depth, flipper length, body mass, and sex columns have missing values.
3. All of the columns seem to be of a reasonable type. 
4. Not any that I can tell of. 

## Task 2: Cleaning Pipeline

Build a step-by-step cleaning pipeline. **Comment every transformation** explaining WHY you made that choice:

```{r}
df_clean <- df |>
  # Step 1: Drop the 2 rows missing ALL four core measurements (bill length/depth,
  # flipper length, body mass). These rows carry zero usable morphometric data —
  # imputing them would mean fabricating values, not cleaning, so removal is the
  # honest choice.
  filter(
    !is.na(bill_length_mm), !is.na(bill_depth_mm),
    !is.na(flipper_length_mm), !is.na(body_mass_g)
  ) |>
  # Step 2: Recode missing `sex` to an explicit "unknown" level instead of dropping
  # those rows. These 9 penguins still have valid measurements — sex just couldn't
  # be determined in the field, so discarding them would throw away good data for
  # a variable we may not even be analyzing.
  mutate(sex = factor(if_else(is.na(sex), "unknown", as.character(sex)))) |>
  # Step 3: Cast `year` to a factor. It's used as a categorical study-year label
  # (2007/2008/2009), not a continuous quantity, so treating it as numeric would
  # invite nonsensical operations like averaging years.
  mutate(year = factor(year))

glimpse(df_clean)
```

## Task 3: Before/After Comparison

Show summary statistics before and after cleaning:

```{r}
library(skimr)

cat("=== BEFORE ===\n")
skim(df)

cat("\n=== AFTER ===\n")
skim(df_clean)
```

**Your turn:** What changed? What improved? Did you lose any data, and was that loss justified?

> We went from 344 rows to 342; only the 2 rows with no morphometric measurements at all were dropped, so that loss is justified since those rows had nothing to analyze. `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, and `body_mass_g` now have zero missing values instead of a few each. `sex` also shows zero `NA`s in the after-summary, but no measurement data was lost for it — the 9 previously-missing values were recoded to an explicit "unknown" category rather than dropped, so those rows are still fully usable for everything except sex-based comparisons. `year` moved from a numeric to a factor summary, which is more appropriate since it's a study-year label, not a quantity to average.

## Task 4: Document Your Decisions

For each cleaning step, complete this table:

| Step | What You Did | Why | What You Considered But Rejected |
|------|-------------|-----|--------------------------------|
| 1 | Dropped the 2 rows missing all four core measurements (bill length/depth, flipper length, body mass) | These rows have zero usable morphometric data, nothing to impute from | Imputing (e.g., with column means): rejected because there's no real information in the row to base an estimate on, so it would fabricate data |
| 2 | Recoded missing `sex` values to an explicit `"unknown"` factor level instead of dropping the rows | Those 9 penguins still have valid measurements; the missingness is in a field-observation variable, not the core data | Dropping rows with missing `sex`: rejected because it would discard good measurement data just to keep one column complete |
| 3 | Cast `year` from numeric to a factor | `year` is a categorical study-year label (2007/2008/2009), not a continuous quantity | Leaving it numeric: rejected because it invites meaningless operations like averaging years |

## Submission Checklist

- [x] Data quality assessment completed
- [x] Cleaning pipeline with comments explaining every step
- [x] Before/after summary statistics shown
- [x] Decision documentation table filled in
- [x] Committed to GitHub

```bash
git add week04/exercises/RP4.2*
git commit -m "Complete Exercise RP4.2: Wrangling challenge with documented decisions"
git push origin main
```
